dispRity manual

### 2017-03-23 doc/dispRity-manual.Rmd
Changed "series" to "subsamples" across the whole package!

-   **Sub-samples**. Sub-samples of the ordinated space.
    A sub-sample (or series) contains the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
    For example, if our ordinated space is composed of birds and mammals (the elements) and 50 principal components (the dimensions), we can create two series containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.

<!-- TG: good call for changing series to subsamples. NEED TO BE APPLIED TO THE WHOLE PACKAGE! -->

@@

```{r, fig.width=6, fig.height=6}
## Splitting the data into time series
sliced_data <- time.series(BeckLee_mat99, BeckLee_tree, method = "continuous",
    model = "acctran", time = rev(seq(from=0, to=130, by=15)),
    FADLAD = BeckLee_ages)

@@

One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter called *series*, see [glossary](#glossary)).
These series correspond to regions of the ordinated space that we want to compare to each other.

@@

`cust.series` # NC: I'd have called this custom.series. It's only 2 more letters and makes it much more readable, and less about custard.
-------------

The `cust.series` function allows to create these series according to one or more factors determined by the user.
In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (all the living mammals and their direct ancestor) and the stem mammals (all the mammals that have no related living descendants).
To do so, we need to separate the tree into two subtrees at the node 71 (see [phylogeny above](#mammal_tree)).

@@

We can then feed these lists of species to the `cust.series` function after transforming it into a factorial table:
<!-- # NC: This is a fairly horrible set up to go through! Can't you make a function? -->

```{r}
## Creating the factor table
series_factors <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
    ncol = 1, dimnames = list(c(crown, stem), "Group")))
head(series_factors)

## Creating the two series
crown_stem <- cust.series(BeckLee_mat50, series_factors)
```

@@

The two series represent two groups occupying specific regions of the mammalian morphospace.
The `dispRity` object itself always contains the original morphospace (`$matrix`) as well as information on which taxa compose the series (`$series`):

```{r}
## This object contains three elements
names(crown_stem)
## With "matrix" being the original matrix (the ordinated space)
str(crown_stem$matrix)
## And "series" the elements from this ordinated space into the two sub-spaces
str(crown_stem$series)
```

@@

```{r}
## Adding all mammals to the a new series (along with crown and stem as before)
series_factors2 <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem)),
    rep("Mammalia", Ntip(BeckLee_tree))), ncol = 2,
    dimnames = list(c(crown, stem), c("Group", "Class")), byrow = FALSE))

## Creating the three series: Group.stem/crown and Class.Mammalia
cust.series(BeckLee_mat50, series_factors2)
```

`time.series`  <a name="time.series"></a>
-------------

Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
The `time.series` function allows to create series containing all the elements present at specific points in time or during specific periods.
This functions needs as input an ordinated space and a matching phylogenetic tree.
Two types of time series can be performed by using the `method` option:

 *  discrete time series (or time-binning) using `method = discrete`;

 *  continuous time series (or time-slicing) using `method = continuous`.

@@

```{r}
## Generating three time bins containing the data present every 40 Ma
time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
    time = c(120, 80, 40, 0))
```

@@

```{r}
## Generating three time bins containing the data present every 40 Ma, including taxa that might
## span between time bins.
(time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages))
```

This generated indeed a `dispRity` object of 3 series.
Note that we can also generate equivalent results by just telling the function that we want three time-bins (series) as follow:

```{r}
## Automatically generate three equal length bins:
time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
    time = 3)
```

@@

```{r}
## Generating four time slices every 40 million years assuming a gradual evolutionary model
(time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages))
## Note that in the same way as before, we can also automatically generate the slices
time.series(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
    model = "gradual", time = 4)
```

@@

Once we obtain our different series, it is possible to bootstrap and rarefy them (i.e. pseudo-replicating the data).
The bootstrap will allow us to make each series more robust to outliers and the rarefaction will allow us to compare series with the same number of elements to get rid of eventual sampling problems.
The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
The default options will bootstrap the matrix 1000 times without rarefaction using the "full" bootstrap method (see below).

@@

This function also allows to rarefy the data using the `rarefaction` argument.
The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e. remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in each series to a minimum of 3 elements).
It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
When left missing, all dimensions from the ordinated matrix are used.

@@

Of course, one could be interested in directly supplying the series generated above directly to this function.
In fact, it can also deal with a list of matrices or with a `dispRity` object output from the `cust.series` or `time.series` functions.

```{r}
## Bootstrapping and fully rarefying the crown/stem series
(crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE))

## Bootstrapping on the time binning/slicing series
(time_binsBS <- boot.matrix(time_bins, bootstraps = 100))
(time_slicesBS <- boot.matrix(time_slices, bootstraps = 100))
```

@@

In these examples we used only simple ordinated matrix but of course, it possible to directly use the `dispRity` objects we generated in the steps above.
Thus we can calculate the bootstrapped and rarefied disparity in each series of the ordinated space.
Here we are going to define disparity as being the median distance between each element and each dimensions' centre.

@@

> Information about the number of elements per series as well as the observed (i.e. non-bootstrapped) disparity is also calculated.
This is specifically handy when rarefying the data for example

@@

> This argument can be left missing. In this case, the algorithm will automatically detect the type of series from the `dispRity` object.

It is also possible to display the number of elements per series (has a horizontal doted line) using the option `elements = TRUE`.
Additionally, when the data is rarefied, one can also indicate which level of rarefaction to display (i.e. display the results for a certain number of elements) by using the `rarefaction` argument.

@@

```{r, fig.width=6, fig.height=6}
## Graphical options
par(bty = "n")
## Plotting the results with some plot.dispRity arguments
plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)), cent.tend = sd, type = "c",
    elements = TRUE, col = c("black", rainbow(10)), ylab = c("Disparity", "Diversity"),
    xlab = "Time (in in units from past to present)", time.series = FALSE,
    observed = TRUE, main = "Many options...")
```

@@

 *   `pairwise` (default): to compare each series pairwise

 *  `referential`: to compare each series to the first one

 *  `sequential`: to compare each series to the following one

 *  `all`: to compare all the series together (like in analysis of variance)

It is also possible to input a list of pairs of `numeric` values or `characters` matching the series names to create personalised test.
Some other specific tests implemented in `dispRity` such as the `dispRity::null.test` and the `dispRity::sequential.test` have a specific way to be applied to the data and therefore ignore the `comparisons` argument.

@@

### `dispRity` object utilities  <a name="dispRity.utilities"></a>
The first series of utilities are functions for manipulating `dispRity` objects:

@@

```{r}
## Extracting the matrix containing the coordinates for the crown mammals
str(matrix.dispRity(disp_crown_stemBS, "Group.crown"))
## Extracting the 3rd boostrapped matrix with the 10th rarefaction level from the same group
str(matrix.dispRity(disp_crown_stemBS, series = 1, bootstrap = 3, rarefaction = 10))
```

#### `get.series.dispRity`
This function creates a dispRity object that contains only elements from one specific series.

```{r}
## Extracting all the data for the crown mammals
(crown_mammals <- get.series.dispRity(disp_crown_stemBS, "Group.crown"))
## The object keeps the properties of the parent object but is composed of only one series
length(crown_mammals$series)
```

@@

## Extracting the disparity from the bootstrapped values from the 10th rarefaction level from the
## second series (stem mammals)
extract.dispRity(disp_crown_stemBS, observed = FALSE, series = 2, rarefaction = 10)
```

@@

```{r}
## Getting the disparity values of the time series
summary(disp_time_slicesBS)

@@

#### `sort.dispRity`
This is the S3 method of `sort` for sorting the series alphabetically (default) or following a specific pattern.

```{r}
## Sorting the disparity series in inverse alphabetic order
summary(sort(disp_time_slicesBS, decreasing = TRUE))
```

@@

This function is at the core of `time.series` when using the time slicing method.
It allows to slice a phylogenetic tree at a certain time point and selecting the nodes or tips following the models detailed in the `time.series` [section above](#time.series).

>>>>

-   **Subsamples**. Sub-samples of the ordinated space.
    A sub-sample (or subsamples) contains the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
    For example, if our ordinated space is composed of birds and mammals (the elements) and 50 principal components (the dimensions), we can create two subsamples containing just mammals or birds, but with the same 50 dimensions, to compare disparity in the two clades.

@@

```{r, fig.width=6, fig.height=6}
## Splitting the data into time subsamples
sliced_data <- time.subsamples(BeckLee_mat99, BeckLee_tree, method = "continuous",
    model = "acctran", time = rev(seq(from=0, to=130, by=15)),
    FADLAD = BeckLee_ages)

@@

One of the first steps in disparity analysis is to split the ordinated space into sub-samples of this space (hereafter called *subsamples*, see [glossary](#glossary)).
These subsamples correspond to regions of the ordinated space that we want to compare to each other.

@@

`custom.subsamples` # NC: I'd have called this custom.subsamples. It's only 2 more letters and makes it much more readable, and less about custard.
-------------

The `custom.subsamples` function allows to create these subsamples according to one or more factors determined by the user.
In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (all the living mammals and their direct ancestor) and the stem mammals (all the mammals that have no related living descendants).
To do so, we need to separate the tree into two subtrees at the node 71 (see [phylogeny above](#mammal_tree)).

@@

We can then feed these lists of species to the `custom.subsamples` function after transforming it into a factorial table:
<!-- # NC: This is a fairly horrible set up to go through! Can't you make a function? -->

```{r}
## Creating the factor table
subsamples_factors <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
    ncol = 1, dimnames = list(c(crown, stem), "Group")))
head(subsamples_factors)

## Creating the two subsamples
crown_stem <- custom.subsamples(BeckLee_mat50, subsamples_factors)
```

@@

The two subsamples represent two groups occupying specific regions of the mammalian morphospace.
The `dispRity` object itself always contains the original morphospace (`$matrix`) as well as information on which taxa compose the subsamples (`$subsamples`):

```{r}
## This object contains three elements
names(crown_stem)
## With "matrix" being the original matrix (the ordinated space)
str(crown_stem$matrix)
## And "subsamples" the elements from this ordinated space into the two sub-spaces
str(crown_stem$subsamples)
```

@@

```{r}
## Adding all mammals to the a new subsamples (along with crown and stem as before)
subsamples_factors2 <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem)),
    rep("Mammalia", Ntip(BeckLee_tree))), ncol = 2,
    dimnames = list(c(crown, stem), c("Group", "Class")), byrow = FALSE))

## Creating the three subsamples: Group.stem/crown and Class.Mammalia
custom.subsamples(BeckLee_mat50, subsamples_factors2)
```

`time.subsamples`  <a name="time.subsamples"></a>
-------------

Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
The `time.subsamples` function allows to create subsamples containing all the elements present at specific points in time or during specific periods.
This functions needs as input an ordinated space and a matching phylogenetic tree.
Two types of time subsamples can be performed by using the `method` option:

 *  discrete time subsamples (or time-binning) using `method = discrete`;

 *  continuous time subsamples (or time-slicing) using `method = continuous`.

@@

```{r}
## Generating three time bins containing the data present every 40 Ma
time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
    time = c(120, 80, 40, 0))
```

@@

```{r}
## Generating three time bins containing the data present every 40 Ma, including taxa that might
## span between time bins.
(time_bins <- time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree,
    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages))
```

This generated indeed a `dispRity` object of 3 subsamples.
Note that we can also generate equivalent results by just telling the function that we want three time-bins (subsamples) as follow:

```{r}
## Automatically generate three equal length bins:
time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
    time = 3)
```

@@

```{r}
## Generating four time slices every 40 million years assuming a gradual evolutionary model
(time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages))
## Note that in the same way as before, we can also automatically generate the slices
time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
    model = "gradual", time = 4)
```

@@

Once we obtain our different subsamples, it is possible to bootstrap and rarefy them (i.e. pseudo-replicating the data).
The bootstrap will allow us to make each subsamples more robust to outliers and the rarefaction will allow us to compare subsamples with the same number of elements to get rid of eventual sampling problems.
The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
The default options will bootstrap the matrix 1000 times without rarefaction using the "full" bootstrap method (see below).

@@

This function also allows to rarefy the data using the `rarefaction` argument.
The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e. remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in each subsamples to a minimum of 3 elements).
It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
When left missing, all dimensions from the ordinated matrix are used.

@@

Of course, one could be interested in directly supplying the subsamples generated above directly to this function.
In fact, it can also deal with a list of matrices or with a `dispRity` object output from the `custom.subsamples` or `time.subsamples` functions.

```{r}
## Bootstrapping and fully rarefying the crown/stem subsamples
(crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE))

## Bootstrapping on the time binning/slicing subsamples
(time_binsBS <- boot.matrix(time_bins, bootstraps = 100))
(time_slicesBS <- boot.matrix(time_slices, bootstraps = 100))
```

@@

In these examples we used only simple ordinated matrix but of course, it possible to directly use the `dispRity` objects we generated in the steps above.
Thus we can calculate the bootstrapped and rarefied disparity in each subsamples of the ordinated space.
Here we are going to define disparity as being the median distance between each element and each dimensions' centre.

@@

> Information about the number of elements per subsamples as well as the observed (i.e. non-bootstrapped) disparity is also calculated.
This is specifically handy when rarefying the data for example

@@

> This argument can be left missing. In this case, the algorithm will automatically detect the type of subsamples from the `dispRity` object.

It is also possible to display the number of elements per subsamples (has a horizontal doted line) using the option `elements = TRUE`.
Additionally, when the data is rarefied, one can also indicate which level of rarefaction to display (i.e. display the results for a certain number of elements) by using the `rarefaction` argument.

@@

```{r, fig.width=6, fig.height=6}
## Graphical options
par(bty = "n")
## Plotting the results with some plot.dispRity arguments
plot(disp_time_slicesBS, quantile = c(seq(from=10, to=100, by=10)), cent.tend = sd, type = "c",
    elements = TRUE, col = c("black", rainbow(10)), ylab = c("Disparity", "Diversity"),
    xlab = "Time (in in units from past to present)", time.subsamples = FALSE,
    observed = TRUE, main = "Many options...")
```

@@

 *   `pairwise` (default): to compare each subsamples pairwise

 *  `referential`: to compare each subsamples to the first one

 *  `sequential`: to compare each subsamples to the following one

 *  `all`: to compare all the subsamples together (like in analysis of variance)

It is also possible to input a list of pairs of `numeric` values or `characters` matching the subsamples names to create personalised test.
Some other specific tests implemented in `dispRity` such as the `dispRity::null.test` and the `dispRity::sequential.test` have a specific way to be applied to the data and therefore ignore the `comparisons` argument.

@@

### `dispRity` object utilities  <a name="dispRity.utilities"></a>
The first subsamples of utilities are functions for manipulating `dispRity` objects:

@@

```{r}
## Extracting the matrix containing the coordinates for the crown mammals
str(matrix.dispRity(disp_crown_stemBS, "Group.crown"))
## Extracting the 3rd boostrapped matrix with the 10th rarefaction level from the same group
str(matrix.dispRity(disp_crown_stemBS, subsamples = 1, bootstrap = 3, rarefaction = 10))
```

#### `get.subsamples.dispRity`
This function creates a dispRity object that contains only elements from one specific subsamples.

```{r}
## Extracting all the data for the crown mammals
(crown_mammals <- get.subsamples.dispRity(disp_crown_stemBS, "Group.crown"))
## The object keeps the properties of the parent object but is composed of only one subsamples
length(crown_mammals$subsamples)
```

@@

## Extracting the disparity from the bootstrapped values from the 10th rarefaction level from the
## second subsamples (stem mammals)
extract.dispRity(disp_crown_stemBS, observed = FALSE, subsamples = 2, rarefaction = 10)
```

@@

```{r}
## Getting the disparity values of the time subsamples
summary(disp_time_slicesBS)

@@

#### `sort.dispRity`
This is the S3 method of `sort` for sorting the subsamples alphabetically (default) or following a specific pattern.

```{r}
## Sorting the disparity subsamples in inverse alphabetic order
summary(sort(disp_time_slicesBS, decreasing = TRUE))
```

@@

This function is at the core of `time.subsamples` when using the time slicing method.
It allows to slice a phylogenetic tree at a certain time point and selecting the nodes or tips following the models detailed in the `time.subsamples` [section above](#time.subsamples).
